---
title: "IMPACT OF WEATHER ON MARATHON PERFORMANCE ACROSS AGE AND GENDER"
subtitle: "PHP2550 Project 1: An exploratory data analysis"
author: "Yanwei (Iris) Tong"
date: "2024-09-29"
output: 
  pdf_document:
    latex_engine: xelatex
link-citations: yes
header-includes:
   - "\\usepackage{xcolor}"
abstract: 
bibliography: 0-references.bib
---

```{r setup, include=FALSE}
# Set up knit environment
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      error = FALSE)
#knitr::opts_chunk$set(out.width = "300px")
# knitr::write_bib(x = c("rmarkdown", "knitr"), file = "0-references.bib")

# Load necessary packages
library(tidyverse)
library(kableExtra)
library(knitr)
library(ggplot2)
library(naniar)
library(gtsummary)

# Define data path
data_path = "/Users/yanweitong/Documents/PHP2550-Data/Project1"

# Import datasets
main_data = read.csv(paste0(data_path, "/project1.csv"))
aqi_data = read.csv(paste0(data_path, "/aqi_values.csv"))
record_data = read.csv(paste0(data_path, "/course_record.csv"))
```

```{r}
# Merge main and record data sets
record_data <- record_data %>%
  mutate(Sex = ifelse(Gender == "F", 0, 1)) %>%
  mutate(Race_code = case_when(Race == "B"~0,
                               Race == "C"~1,
                               Race == "NY"~2,
                               Race == "TC"~3,
                               Race == "D" ~4))

merged_main <- main_data %>%
  left_join(record_data[, c("Year", "Sex", "CR", "Race_code")], 
            by = c("Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D." = "Race_code", 
                   "Year" = "Year",
                   "Sex..0.F..1.M." = "Sex")) %>%
  dplyr::rename(Race_code = Race..0.Boston..1.Chicago..2.NYC..3.TC..4.D.,
         Sex = Sex..0.F..1.M.,
         Age = Age..yr.) %>%
  mutate(Gender = factor(ifelse(Sex == 0, "Female", "Male"))) %>%
  mutate(Marathon = case_when(Race_code == 0 ~ "Boston",
                          Race_code == 1~ "NYC",
                          Race_code == 2 ~ "Chicago",
                          Race_code == 3 ~ "Twin Cities",
                          Race_code == 4 ~ "Grandmas")) 


#Clean up AQI by CBSA
aqi_data = aqi_data %>% 
  distinct()

aqi_mean = aqi_data %>%
  group_by(marathon, date_local, parameter, sample_duration) %>%
  summarise(daily_mean = mean(arithmetic_mean, na.rm = TRUE)) %>%
  mutate(parameter_duration = paste0(parameter, "-", sample_duration)) 

aqi_pivot = aqi_mean[,c("marathon", "date_local", "parameter_duration", "daily_mean")] %>%
  pivot_wider(names_from = parameter_duration, values_from = daily_mean) %>%
  mutate(Year = year(date_local))

merged_main = merged_main %>% 
  left_join(aqi_pivot, 
            by = c("Marathon" = "marathon",
                   "Year" = "Year"))
```


## \textcolor{orange}{INTRODUCTION}
Here, we would like to identify and explore the impacts of the following factors on marathon performance across genders: 1) the increase in age, 2) environmental conditions and whether the impact differs across age and gender, and 3) 

## \textcolor{orange}{METHODS}
### Data Overview
```{r}
# Baseline summary
merged_main %>%
  select(
    Marathon,
    Gender, 
    Age
  ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    by = Marathon,
    digits = all_continuous() ~ 2,
    missing = "no",
    type = list(
      Gender ~ "categorical"
    )
  ) %>%
  add_p() %>%
  modify_caption(caption = "Participant Characteristics by Marathon, N = {N}") %>%
  as_kable_extra(
    booktabs = TRUE,
    longtable = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    position = "center",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )

```

```{r}
merged_main  %>%
  dplyr::select(
    Marathon,
    WBGT,
    DP,
    `Nitrogen dioxide (NO2)-1 HOUR`,
    `Sulfur dioxide-1 HOUR`,
    `Ozone-1 HOUR`,
    `PM2.5 - Local Conditions-1 HOUR`
  ) %>%
  distinct() %>%
  tbl_summary(
    statistic = all_continuous() ~ c("{mean} ({sd})", "{min}, {max}"),
    by = Marathon,
    digits = all_continuous() ~ 2,
    missing = "no",
    type = list(
      WBGT ~ "continuous2",
      DP ~"continuous2",
      `Sulfur dioxide-1 HOUR`  ~"continuous2",
      `Nitrogen dioxide (NO2)-1 HOUR` ~"continuous2",
      `Ozone-1 HOUR` ~"continuous2",
      `PM2.5 - Local Conditions-1 HOUR` ~"continuous2"
    ),
    label = list(
      DP = "DP (°C)",
      `Nitrogen dioxide (NO2)-1 HOUR` = "NO$_2$ (parts per billion)",
      `Sulfur dioxide-1 HOUR` = "SO$_2$ (parts per billion)",
      `Ozone-1 HOUR` = "O$_3$, (parts per million)",
      `PM2.5 - Local Conditions-1 HOUR` = "PM$_{2.5}, (μg/m$^3$)$"
      # ,
      # Mat_race = "Maternal race",
      # SGA = "Small for gestational age",
      # gender = "Infant gender",
      # ga = "Obstetrical gestational age (weeks)",
      # Del_method = "Delivery method",
      # mat_chorio = "Maternal Chorioamnionitis"
    )
  ) %>%
  modify_caption(caption = "Summary of weather and pollution parameters across five marathons, N = {N}") %>%
  as_kable_extra(
    booktabs = TRUE,
    longtable = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    position = "center",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )

```



### Statistical Analysis

## \textcolor{orange}{RESULTS}

```{r, fig.width = 4, fig.height = 3}
best_time_gender_age = ggplot(merged_main, aes(x = Age, y = X.CR, color = Gender)) +
   geom_smooth(method = "loess", se = TRUE, size = 1) +   # Loess smoothing with 95% CI
  labs(
    title = "Men vs Women",
    x = "Age (yrs)",
    y = "Best Time (%CR)"
  ) +
  scale_y_continuous(limits = c(0, 300)) +  
   scale_color_manual(values = c("Male" = "steelblue", "Female" = "darkred")) + 
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    legend.title = element_blank(), 
    legend.position = "right" 
  )
best_time_gender_age
```



## \textcolor{orange}{CONCLUSION}


## \textcolor{orange}{DISCUSSION}


\pagebreak
## Data Privacy

## Code availability

## Reference


\pagebreak
## **Code Appendix**
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


